OpenMPIR

### 2017-09-08 paper.tex
Changed the discussion of optimization vs runtime

When writing task-parallel programs today, one has a large selection of
programming models. Unfortunately, despite having significant overlap in
semantics, parallel programming models such as OpenMP\cite{}, Cilk\cite{},
Kokkos\cite{}, HPX\cite{}, Charm++\cite{},
Qthreads\cite{}, pthreads\cite{}, MPI\cite{}, Chapel\cite{}, UPC\cite{},
among others have virtually no ability to interoperate either at compile
time or run time.
\wmnote{Not sure if relevant, but there was some experiment run a while back of someone composing a library that internally used OpenMP
with another framework that had terribly results compared with either used alone.}
Many of these programming models are implemented as runtime libraries with only
primitive compiler support. As a result, most of these parallel frameworks
aren't able to take advantage of the compiler's analysis and optimizations.
There has been some recent work on specializing compilers to reason
about parallel programming libraries \cite{Moss_2016}, but fragmentation among
models presents a significant challenge for compiler writers, which must choose
a single model to analyze and optimize. An ideal solution would be some form of
common representation of parallelism.

@@

\gsnote{I wanted to give a brief nod to the importance of LLVM, given the venue
(LLVM in HPC). Is this any more satisfactory?}
\wmnote{It still feels like we're trying to advocate for LLVM. I'd prefer some
more objective-sounding language like perhaps the one below.}
LLVM has gained traction in the academic community for its simplicity,
relatively painless modification, and extensibility\cite{lattner2004llvm}.
Historically, LLVM has had a sequential semantics: there is no explicit notion
of concurrent or parallel processes. This has resulted in existing compilers
treating parallelism constructs as thin wrappers for calls into runtime
systems. These calls are infeasible for the compiler to reason about, due to
the complexity and flexibility of runtimes being called. This situation was
recently remedied by Schardl et al. \cite{tapir}. By adding three instructions
to LLVM, Tapir enables analyses and optimizations of parallel code. We describe
Tapir in detail in Section~\ref{Sec:Tapir}. For a more complete description see
\cite{tapir}.

@@

If one were to compile this program using the traditional technique of converting
to runtime calls, one would find a program similar to \figref{runtime_calls}.
The OpenMP pragmas are effectively treated as syntactic sugar for such runtime calls
because vanilla LLVM has no concise way to represent parallel programs.
These runtime calls obfuscate the program, making it impossible to analyze or
optimize parallel code.

@@

\caption[Simplified compilation of the unoptimized search code from \figref{search}.]{Simplified compilation of the unoptimized search code from \figref{search}.  \subfigcap{a}~The function \code{search}. The parallel tasks are moved into their own functions and accompanying closures. These functions are then passed to OpenMP runtime calls. This system obfuscates the program, making it impossible for an optimizer to reason about what is happening.}
  \label{fig:runtime_calls}
\end{figure}

@@

While we would like to distinguish between these, it is difficult to do so with
only performance counter information. A more thorough investigation would
require deep knowledge of each of the frontends and runtimes. While the
importance of optimizations of parallel code is clear, we hypothesize that most
of the disparity present in these results is a function of runtime
implementation differences. Our evidence for this comes from Schardl et al.
\cite{tapir}. Generally, the effect of the optimizations that have been
implemented for Tapir IR are outweighed by runtime contention, memory accesses,
etc. for parallel runs. \gsnote{Billy, this OK? Don't want to sound too
negative about the importance of the optimizations you've implemented.}
\wmnote{So actually the memory accesses are happily an optimization Tapir does
-- mainly some mem2reg optimization that may not make much of a different (though still something)
on one core, is a lot more impactful on multicores because of issues regarding contention
forcing a cache line flush.
Also I'm not convinced there wasn't an impact on FFT/nqueens because we did see a
decent improvement from tapir on that -- though I wouldn't be suprised to see the runtime
contribute a large chunk of that.
Fibonacci is certainly runtime based because there's basically no work inside of
it so its almost entirely a measurement of the runtime efficiency.
Sort I'm not sure on depending on the implementation
}

>>>>

When writing task-parallel programs today, one has a large selection of
programming models. Unfortunately, despite having significant overlap in
semantics, parallel programming models such as OpenMP\cite{}, Cilk\cite{},
Kokkos\cite{}, HPX\cite{}, Charm++\cite{},
Qthreads\cite{}, pthreads\cite{}, MPI\cite{}, Chapel\cite{}, UPC\cite{},
among others have virtually no ability to interoperate either at compile
time or run time.
\wmnote{Not sure if relevant, but there was some experiment run a while back of someone composing a library that internally used OpenMP
with another framework that had terribly results compared with either used alone.} 
\gsnote{Hmm. A quick googling turned up a project out of berkeley combining OpenMP and TBB, 
but performance didn't seem too bad. I'll take another look tomorrow.}
Many of these programming models are implemented as runtime libraries with only
primitive compiler support. As a result, most of these parallel frameworks
aren't able to take advantage of the compiler's analysis and optimizations.
There has been some recent work on specializing compilers to reason
about parallel programming libraries \cite{Moss_2016}, but fragmentation among
models presents a significant challenge for compiler writers, which must choose
a single model to analyze and optimize. An ideal solution would be some form of
common representation of parallelism.

@@

LLVM has gained traction in the academic community for its simplicity,
relatively painless modification, and extensibility\cite{lattner2004llvm}.
Historically, LLVM has had a sequential semantics: there is no explicit notion
of concurrent or parallel processes. This has resulted in existing compilers
treating parallelism constructs as thin wrappers for calls into runtime
systems. These calls are infeasible for the compiler to reason about, due to
the complexity and flexibility of runtimes being called. This situation was
recently remedied by Schardl et al. \cite{tapir}. By adding three instructions
to LLVM, Tapir enables analyses and optimizations of parallel code. We describe
Tapir in detail in Section~\ref{Sec:Tapir}. For a more complete description see
\cite{tapir}.

@@

If one were to compile this program using the traditional technique of converting
to runtime calls, one would find a program similar to \figref{runtime_calls}.
The OpenMP pragmas are effectively treated as syntactic sugar for such runtime calls
because vanilla LLVM has no concise way to represent parallel programs.
These runtime calls obfuscate the program, making it infeasable to analyze or
optimize parallel code.

@@

\caption[Simplified compilation of the unoptimized search code from \figref{search}.]{Simplified compilation of the unoptimized search code from \figref{search}.  \subfigcap{a}~The function \code{search}. The parallel tasks are moved into their own functions and accompanying closures. These functions are then passed to OpenMP runtime calls. This system obfuscates the program, making it infeasable for an optimizer to reason about what is happening.}
  \label{fig:runtime_calls}
\end{figure}

@@

While we would like to distinguish between these, it is difficult to do so with
only performance counter information. A more thorough investigation would
require deep knowledge of each of the frontends and runtimes. While the
importance of optimizations of parallel code is clear, we hypothesize that most
of the disparity present in these results is a function of runtime
implementation differences. As justification for this hypothesis, consider
the \texttt{Fibonacci} benchmark. There are no memory accesses to the heap, 
and the code leaves little room for optimization, algorithmic changes
notwithstanding. In other words, it is effectively a benchmark of how fast a
runtime can execute fork-join parallelism. \gsnote{still worth looking into the
optimization I mention in the implemtnation section.}

%\wmnote{So actually the memory accesses are happily an optimization Tapir does
%-- mainly some mem2reg optimization that may not make much of a different
%(though still something) on one core, is a lot more impactful on multicores
%because of issues regarding contention forcing a cache line flush.  Also I'm
%not convinced there wasn't an impact on FFT/nqueens because we did see a decent
%improvement from tapir on that -- though I wouldn't be suprised to see the
%runtime contribute a large chunk of that, especially for nqueens because that
%seems like a much larger improvement than the serial optimizations applied to
%parallel program alone (when we tested).  Fibonacci is certainly runtime based
%because there's basically no work inside of it so its almost entirely a
%measurement of the runtime efficiency.  Sort I'm not sure on depending on the
%implementation
%}
